---
title: "Figures for the Trained immunity project"
author: "Jiangyan Yu (jiangyan.yu@uni-bonn.de)"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: yes
    theme: united
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# libraries & functions

```{r,include=FALSE}
# rm(list = ls())
set.seed(123)

list.cran.packages = c("ggplot2","Seurat","dplyr","magrittr","tidyr",'ggpubr',"Seurat")
new.cran.packages <- list.cran.packages[!(list.cran.packages %in% installed.packages()[,"Package"])]
if(length(new.cran.packages)>0) install.packages(new.cran.packages)

list.bioc.packages = c("nichenetr","monocle3","monocle","SeuratWrappers","edgeR","ComplexHeatmap","AUCell")
new.bioc.packages <- list.bioc.packages[!(list.bioc.packages %in% installed.packages()[,"Package"])]
if(length(new.bioc.packages)>0)if (!requireNamespace("BiocManager")) install.packages("BiocManager")
BiocManager::install(new.bioc.packages, update = FALSE)

lapply(c(list.cran.packages,list.bioc.packages), require, character.only = TRUE)

# rm("list.bioc.packages","list.cran.packages","new.bioc.packages","new.cran.packages" )
```

# working directory

```{r}
# system_dir = "/Users/jiangyanyu/sciebo/"
# system_dir = "/home/yu.j/sciebo/"
system_dir = "/home/jyu/rstudio/"
working_dir = paste0(system_dir,"/Projects/Katzmarski_mouse_macro_ownerJY/")
published_data_dir = paste0(system_dir,"/general_scripts/publised_datasets/")
global_ref_dir =paste0(system_dir,"/general_scripts/Global_ref_data/")
gsea_pathway_dir = paste0(system_dir,"/general_scripts/Global_ref_data/")

source(paste0(global_ref_dir,"general_functions.R"))
```

# process scRNA data

## load saved rdata

following filtering criteria was applied in the file: Schlitzer_seuratObject_firstExperiment_2020-02-07.RData
subset = nFeature_RNA > 500 & nFeature_RNA < 3000 & percent.mt < 10 & percent.Hb < 0.0001

```{r}
# load(paste(working_dir,"Schlitzer_seuratObject_firstExperiment_2020-02-07.RData",sep=""))
# 
# run01_counts = seurat@assays$RNA@counts
# jonas_run01_metadata = seurat@meta.data
# 
# run01_seurat = CreateSeuratObject(run01_counts, min.cells = 3, project = "Katzmarski_seurat_run01_allcells", meta.data = seurat@meta.data[,c("sample","status")])
# rm(run01_counts,seurat)
# 
# run01_seurat[["percent.mt"]] <- PercentageFeatureSet(run01_seurat, pattern = "^mt-")
# FeatureScatter(run01_seurat,feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# VlnPlot(run01_seurat,features = c("nCount_RNA","nFeature_RNA","percent.mt"))
```

## re-run seurat

```{r}
# run01_seurat = run01_seurat %>% NormalizeData(., normalization.method = "LogNormalize", scale.factor = 10000) %>% FindVariableFeatures(., selection.method = "vst", nfeatures = 2000) %>% ScaleData(., features = rownames(.)) %>% RunPCA(.,features = VariableFeatures(.), verbose = FALSE) %>% JackStraw(., num.replicate = 100) %>% ScoreJackStraw(., dims = 1:20)
# 
# JackStrawPlot(run01_seurat, dims = 1:20)
# ElbowPlot(run01_seurat)
# 
# run01_seurat = FindNeighbors(run01_seurat, dims = 1:10)
# 
# # set resolution as 0.25 to generate 6 clusters, as same as the output from time series experiment
# run01_seurat = run01_seurat %>% FindClusters(., resolution = 0.25) %>% RunUMAP(., dims=1:10)
# 
# 
# #UMAP
# run01_seurat <- RunUMAP(run01_seurat,
#                      dims = 1:30,
#                      n.neighbors = 30,
#                      min.dist = 1,
#                      n.components = 3,
#                      spread = 3,
#                      a=1,
#                      b=0.8)

```

## save/load reanalyzed file

```{r}
# saveRDS(run01_seurat,paste(working_dir,"JYu_Run01_reanalysis_20220517.rds"))

run01_seurat = readRDS(paste(working_dir,"JYu_Run01_reanalysis_20220517.rds"))
```

# generate figures

##1.Combined UMAP of D7 PBS vs. Glucan

```{r}
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"

cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")


# pdf(file = paste0(ms_folder,"1.day7_merged_umap.pdf"),width = 5, height = 4,colormodel="cmyk")
DimPlot(run01_seurat,dims = c(1,2),pt.size=0.3)+
  scale_color_manual(values = cluster_color)+
  theme_classic()+
  labs(subtitle = "Run01_day7_PBSandGLUCAN (10202 cells)")
# dev.off()

rm(ms_folder,cluster_color)
```

##2.UMAP of D7 PBS

```{r}
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"

cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")


# pdf(file = paste0(ms_folder,"2.day7_pbs_umap.pdf"),width = 5, height = 4,colormodel="cmyk")
DimPlot(subset(run01_seurat,subset=status=="Solvent"),dims = c(1,2),pt.size=0.3)+
  scale_color_manual(values = cluster_color)+
  theme_classic()+
  labs(subtitle = "Run01_day7_PBS (4845 cells)")
# dev.off()

rm(ms_folder,cluster_color)
```

##3.UMAP of D7 PBS

```{r}
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"

cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")


# pdf(file = paste0(ms_folder,"3.day7_glucan_umap.pdf"),width = 5, height = 4,colormodel="cmyk")
DimPlot(subset(run01_seurat,subset=status=="Glucan"),dims = c(1,2),pt.size=0.3)+
  scale_color_manual(values = cluster_color)+
  theme_classic()+
  labs(subtitle = "Run01_day7_GLUCAN (5357 cells)")
# dev.off()

rm(ms_folder,cluster_color)
```

##4.barplot by cluster

```{r}
ns <- table(organ = run01_seurat$status, cell_type = run01_seurat$seurat_clusters)
## remove cluster 1, 5, 6
# ns = ns[,c(1,3,4,5,8)]
fq <- prop.table(ns, 1) * 100
df <- as.data.frame(fq)

ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")

df$organ = factor(df$organ,levels = c("Glucan","Solvent"))
library(ggplot2)

# pdf(file = paste0(ms_folder,"4.barplot_by_cluster.pdf"),width = 5, height = 2,colormodel="cmyk")


ggplot(df,aes(x=Freq,y=organ,fill=cell_type,label=cell_type))+
  geom_bar(stat="identity",colour="black")+
  scale_fill_manual(values=cluster_color)+
  theme_classic()+
  # geom_text(position = position_stack(vjust = 0.5))+
  theme(axis.title=element_blank())

# dev.off()
rm(ns,fq,df,cluster_color,ms_folder)
```

##5.barplot by cluster

```{r}
ns <- table(organ = run01_seurat$seurat_clusters, cell_type = run01_seurat$status)
## remove cluster 1, 5, 6
# ns = ns[,c(1,3,4,5,8)]
fq <- prop.table(ns, 1) * 100
df <- as.data.frame(fq)

ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
cluster_color = c("white","#A35884")

# df$organ = factor(df$organ,levels = c("Glucan","Solvent"))
library(ggplot2)

# pdf(file = paste0(ms_folder,"5.barplot_by_condition.pdf"),width = 5, height = 2,colormodel="cmyk")


ggplot(df,aes(x=Freq,y=organ,fill=cell_type,label=cell_type))+
  geom_bar(stat="identity",colour="black")+
  scale_fill_manual(values=cluster_color)+
  theme_classic()+
  # geom_text(position = position_stack(vjust = 0.5))+
  theme(axis.title=element_blank())
 
# dev.off()
rm(ns,fq,df,cluster_color,ms_folder)
```
##5.1 APOE violin

```{r}
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
# pdf(file = paste0(ms_folder,"5.1.ApoeViolin.pdf"),width = 4, height = 4,colormodel="cmyk")
cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")
VlnPlot(run01_seurat,features = c("Apoe"),group.by = "seurat_clusters")+scale_fill_manual(values = cluster_color)

# dev.off()
rm(cluster_color,ms_folder)
```

##6.DEG Heatmap D7

```{r}
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")
# pdf(file = paste0(ms_folder,"6.DEG_heatmap_run01.pdf"),width = 6,height = 8)
run01_seurat_deg = FindAllMarkers(run01_seurat,only.pos = FALSE, min.pct = 0.25, logfc.threshold = 0.25)
DoHeatmap(run01_seurat, 
          features = run01_seurat_deg %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC) %>% .$gene,
          group.colors = cluster_color)+
   scale_fill_gradientn(colors = c("blue", "white", "red"))
# dev.off()
```

##7.UMAP Glucan D0, 7, 21

```{r}
# ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
# 
# cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C","white")
# cell_selected = run02_seurat@meta.data[unlist(tapply(1:nrow(run02_seurat@meta.data),run02_seurat@meta.data$combined_condition,function(x) sample(x,978))),]
# cell_selected = subset(cell_selected,combined_condition!="BAL_Glucan_d14")
# 
# # pdf(file = paste0(ms_folder,"7.UMAP_D0_D7_D21_run02.pdf"),width = 12, height = 4,colormodel="cmyk")
# DimPlot(run02_seurat[,rownames(cell_selected)],dims = c(2,3),pt.size=0.3,split.by = "combined_condition")+
#   scale_color_manual(values = cluster_color)+
#   theme_classic()+
#   labs(subtitle = "Run02_glucan (978 cells per condition)")
# # dev.off()
# 
# rm(ms_folder,cluster_color)
```

##8.barplot of gene expressions

SiglecF, ITGAX,, Fapb4, Ear2, Prdx1, Top2a, Mki67, Gpnmb, TREM2, ApoE, ifit2, ifit3, Gbp2

```{r}
tmp_genes = c("Siglecf", "Itgax", "Fabp4", "Ear2", "Prdx1", "Top2a", "Mki67", "Gpnmb", "Trem2", "Apoe", "Ifit2", "Ifit3", "Gbp2")
cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C","white")
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"

# pdf(file = paste0(ms_folder,"8.barplot_of_selected_genes.pdf"),width = 6, height = 2,colormodel="cmyk")
# 
# for(i in c(1:length(tmp_genes))){
#   tmp_data = cbind(run01_seurat@assays$RNA@data[tmp_genes[i],],run01_seurat@meta.data)
#   colnames(tmp_data)[1] = tmp_genes[i]
#   tmp_data = tmp_data[order(tmp_data$seurat_clusters),]
#   
#   p = ggplot(tmp_data,aes(x=c(1:nrow(tmp_data)),y=get(tmp_genes[i]),fill=seurat_clusters))+
#     geom_col(width = 1)+
#     scale_fill_manual(values  = cluster_color)+
#     theme_classic()+
#     theme(axis.title.x=element_blank(),axis.text.x = element_blank(),axis.ticks.x = element_blank())+
#     labs(title = tmp_genes[i],y="Normazlied UMI counts")
#     
#   print(p)
# }
# 
# dev.off()

## save one figure per file

for(i in c(1:length(tmp_genes))){
  # pdf(file = paste0(ms_folder,"8.barplot_of_",tmp_genes[i], ".pdf"),width = 6, height = 2,colormodel="cmyk")
  tmp_data = cbind(run01_seurat@assays$RNA@data[tmp_genes[i],],run01_seurat@meta.data)
  colnames(tmp_data)[1] = tmp_genes[i]
  tmp_data = tmp_data[order(tmp_data$seurat_clusters),]
  
  p = ggplot(tmp_data,aes(x=c(1:nrow(tmp_data)),y=get(tmp_genes[i]),fill=seurat_clusters))+
    geom_col(width = 1)+
    scale_fill_manual(values  = cluster_color)+
    theme_classic()+
    theme(axis.title.x=element_blank(),axis.text.x = element_blank(),axis.ticks.x = element_blank())+
    labs(title = tmp_genes[i],y="Normazlied UMI counts")
    
  print(p)
  # dev.off()
}

rm(p,tmp_genes, cluster_color, ms_folder, tmp_data,i)



```

##8.1.violin of gene expressions

SiglecF, ITGAX,, Fapb4, Ear2, Prdx1, Top2a, Mki67, Gpnmb, TREM2, ApoE, ifit2, ifit3, Gbp2

```{r}
tmp_genes = c("Siglecf", "Itgax", "Fabp4", "Top2a", "Mki67", "Gpnmb", "Apoe", "Ifit2", "Gbp2")
cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C","white")
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"

# pdf(file = paste0(ms_folder,"8.1.vlnplot_of_selected_genes.pdf"),width = 4, height = 4,colormodel="cmyk")

for(i in c(1:length(tmp_genes))){

  p = VlnPlot(run01_seurat,features = tmp_genes[i],group.by = "seurat_clusters")+scale_fill_manual(values = cluster_color)


  print(p)
}

# dev.off()




rm(p,tmp_genes, cluster_color, ms_folder, tmp_data,i)



```

##10.run02_UMAP

```{r}
# ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
# 
# cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")
# 
# 
# # pdf(file = paste0(ms_folder,"10.run02_merged_umap.pdf"),width = 5, height = 4,colormodel="cmyk")
# tmp_seurat= subset(run02_seurat,subset=seurat_clusters!=5) %>% subset(.,subset=combined_condition!="BAL_Glucan_d14")
# 
# DimPlot(tmp_seurat,dims = c(2,3),pt.size=0.3)+
#   scale_color_manual(values = cluster_color)+
#   theme_classic()+
#   labs(subtitle = "Run02_BAL_glucan_D0-D7-D21 (3069 cells)")
# # dev.off()
# 
# rm(ms_folder,cluster_color, tmp_seurat)
```

##11.DEG Heatmap run02

```{r}
# ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
# cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")
# # pdf(file = paste0(ms_folder,"11.DEG_heatmap_run02.pdf"),width = 6,height = 8)
# tmp_seurat= subset(run02_seurat,subset=seurat_clusters!=5) %>% subset(.,subset=combined_condition!="BAL_Glucan_d14")
# 
# 
# DoHeatmap(tmp_seurat, 
#           features = run02_seurat_deg %>% subset(.,cluster!=5) %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC) %>% .$gene,
#           group.colors = cluster_color)+
#    scale_fill_gradientn(colors = c("blue", "white", "red"))
# # dev.off()
# 
# rm(ms_folder,cluster_color, tmp_seurat)
```

##12.QC plots

```{r}
ms_folder = "/home/yu.j/sciebo/Projects/Katzmarski_mouse_macro_ownerJY/ms_figures_reanalysis20220517/"
cluster_color = c("#3A3A3C","#A9DDD9","#0E7D63","#A35884","#EEBA4C")
# pdf(file = paste0(ms_folder,"12.1.QC_for_run01.pdf"),width = 12,height = 4)

VlnPlot(run01_seurat,features = c("nCount_RNA","nFeature_RNA","percent.mt"),group.by = "seurat_clusters")+scale_fill_manual(values = cluster_color)

# dev.off()



```

# session information

```{r}
sessionInfo()
```

